---
title: |
  | Data Management II 
  | using dplyr and forcats
author: |
  | Kamarul Imran Musa
  | Assoc Prof (Epidemiology and Statistics)
date: '`r Sys.Date()`'
output: 
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Data transformation (data munging or data wrangling)

## Definition of data wrangling 

1.  Data munging or data wrangling is loosely the process of manually converting or mapping data from one "raw" form into another format.
2.  The process allows for more convenient consumption of the data. 
3.  In doing so, we will be using semi-automated tools. 

Refer <https://community.modeanalytics.com/sql/tutorial/data-wrangling-with-sql/>

## Package: dplyr 

### About dplyr package 

**dplyr** is a package grouped inside **tidyverse** collection of packages. 

**dplyr** package is a very useful package to munge or wrangle or to tranform your data. 

It is really a grammar of data manipulation. It provides a consistent set of verbs that help you solve the most common data manipulation challenges

Refer <https://github.com/tidyverse/dplyr>

## Data wrangling (or transformation) using dplyr

When we communicate with data, common procedures include

1.  reducing the size of dataset by selecting certain variables (or columns)
2.  generating new variables from existing variables 
3.  sorting observation of a variable 
4.  grouping observations that fulfil certain criteria
5.  reducing variable to groups to in order to estimate summary statistic 

# Using dplyr package 

For the procedures listed above, the corresponding **dplyr** functions are

1.  `select` - to reduce the size of dataset by selecting certain variables  
2.  `mutate` - to generate new variables from existing variables 
3.  `arrange` - to sort observation of a variable
4.  `filter` - to group observations that fulfil certain criteria 
5.  `group_by` and `summarize` - to reduce variable to groups to in order to estimate summary statistic  

## Hands-on 1: Preparation and data

### Working directory and data format

It is very important to ensure you know where your working directory is. 

Every time you want to start processing your data, please make sure:

1.  you have set your working directory. Type `setwd()` will display the active **working directory**.  
2.  then, read (import) the data you want to analyze. 

Remember, there are a number of ways and packages useful to read the data. They depend on the format of your data. For example SPSS format, Stata format, SAS format, MS Excel format and `.csv` format. 

Two useful packages - **haven** and **foreign** packages - are very useful to read or import your data into R memory. Take note that, different packages require you to specify different parameters inside the arguments. 

### Training data

To make life easier and to facilitate reproducibility, we will use examples available from the public domains.

To reproduce the outputs demonstrated on **tidyverse** website <https://github.com/tidyverse/dplyr>, we will use similar dataset or datasets with them. 

One of the useful datasets is 'starwars'. The `starwars` data comes with **dplyr** package. This data comes from SWAPI, the Star Wars API accessible at http://swapi.co/. 

The 'starwars' data is class of `tibble`. The data have:

- 87 rows (observations)  
- 13 columns (variables)

Now, let us:

1.  load the **dplyr** package
2.  examine the column names (variable names) 

```{r}
library(dplyr)
names(starwars)
```

Next, examine the first 10 observations of the data. There are 77 more rows and 7 more variables NOT SHOWN. You can also see the types of the variables `chr`, `int`, `dbl`

```{r}
starwars
```

## Hands-on 2: `select()` and `mutate()` 

### `select()`

When you work with large datasets with many columns, sometimes it is easier to select only necessary columns to create a smaller dataset that you can work on. 

To create smaller datasets, select only some of the columns. This will greatly help data exploration or analysis.  

In `starwars` data, we have 13 variables. Let us select only variables below, then generate a new dataset named as `mysw` :

1.  name
2.  height (cm)
3.  mass (kg) 
4.  gender

```{r}
mysw <- select(starwars, name, gender, height, mass)
mysw
```



### `mutate()`

With `mutate()`, you can generate new variable or variables. 

For example, in the dataset `mysw`, we want to create a new variable `bmi` which equals mass in kg divided by squared height (in meter) $$bmi = \frac{kg}{m^2}$$


```{r}
mysw <- mutate(mysw, bmi = mass/(height/100)^2)
mysw
```

## Hands-on 3: `arrange()` and `filter()`

### arrange

To easily sort (ascending or descending) data, the `arrange` function is useful. 

The `arrange` function will sort the observation based on the values of the specified variable. 

Let create a new dataset (`mysw`) by sorting the observation of variable bmi from the biggest bmi to the lowest bmi:

```{r}
mysw <- arrange(mysw, desc(bmi))
mysw
```

Now, we will replace the dataset `mysw` with data that contain bmi values from the lowest to the biggest bmi 

```{r}
mysw <- arrange(mysw, bmi)
mysw
```


### filter

To group observations based on certain criteria, we use the `filter` function. 

Here, We will create a new dataset `mysw_m_40` that contains only **male** gender and **bmi at or above 30** 

```{r}
mysw_m_40 <- filter(mysw, gender == 'male', bmi >= 30)
mysw_m_40
```

There are missing (NA) bmi observation in the dataset. How about, we create a new dataset containing height above 200 or BMI above 45, BUT does not include NA observation for bmi

```{r}
mysw_ht_45 <- filter(mysw, height >200 | bmi >45, bmi != 'NA')
mysw_ht_45
```

## Hands-on 4: group_by 

### Summarize data - summarize

To use `summarize` function, the `group_by` function is almost always necessary. 

The `group_by` function will prepare the data for group analysis. For example, 

1.  to get summary values for mean **bmi**, mean **ht** and mean **mass**
2.  for male, female, hermaphrodite and none (**gender** variable)

```{r}
mysw_g <- group_by(mysw, gender)
summarise(mysw_g, meanbmi = mean(bmi, na.rm = TRUE),
                  meanht  = mean(height, na.rm = TRUE),
                  meanmass = mean(mass, na.rm = TRUE))
```

## Summary

'dplyr' package is a very useful package that encourage users to use proper verb when manipulating variables (columns) and observations (rows). 

We have learned to use 5 functions but there are more functions available. 

Other useful functions are:

1.  distinct()
2.  mutate() and transmute()
3.  sample_n() and sample_frac()

Also note that, package 'dplyr' is very useful when it is combined with another function that is 'group_by'

# Using forcats package

This package helps to work with factor variables. 


## Hands-on 5: forcats

### New dataset

To start with let us create a dummy dataset:

1.  a vector data **sex1** , values = 0,1
2.  a vector data **race1** , values = 1,2,3,4
3.  a dataframe (dataset) **data_f**

```{r}
sex1 <- rbinom(n = 100, size = 1, prob = 0.5) 
str(sex1)
race1 <- rep(seq(1:4), 25)
str(race1)
data_f <- data.frame(sex1, race1)
head(data_f)
```

We can see the data now, above. 

Now let us see the structure of all variables. You should see that they are all in the integer (numerical) format

```{r}
str(data_f)
```

### Convert numeric to factor variables



1.  Generate male variable from sex: sex1 (int) to male (a factor variable). Label as No or Yes 
2.  Generate race2 variable from race1: race1 (int) to race2 (a factor variable). Label as Mal, Chi, Ind, Others

```{r}
data_f$male <- factor(data_f$sex1, labels = c('No', 'Yes')) 
data_f$race2 <- factor(data_f$race1, labels = c('Mal', 'Chi', 'Ind', 'Others')) 
str(data_f)
```

### Recode old to new levels

Steps:

1.  For variable **male**, change from `No` vs `Yes` TO `Fem` and `Male`
2.  Create a new variable **malay** from variable **race2** AND 
3.  Label `Chi` to `Non-Malay`, `Ind` to `Non-Malay` and `Others` to `Non-Malay`. But we keep `Mal` as it is

Now, we need **dplyr** and **forcats** perform steps 1 to 3 above. 

```{r} 
library(dplyr)
library(forcats)
data_f$male <- data_f$male %>% fct_recode('Fem' = 'No', 'Male' = 'Yes')
data_f <- data_f %>% mutate(malay = fct_recode(race2, 
                                               'Non-Malay' = 'Chi', 
                                               'Non-Malay' = 'Ind', 
                                               'Non-Malay' = 'Others'))
head(data_f)
```

# Session 

```{r}
sessionInfo()
```



