---
output:
  pdf_document: default
  html_document:
    keep_md: yes
---

# Introduction to R
This chapter introduces readers to the basics of working with data in R. We will start with installing R in your computer and getting familiar with RStudio interface. These will be followed by the basics of handling data in R.

## R and RStudio

### Installation of R
- The latest version of R is `r version[13:14]`.
- R is available for Windows, Mac OS and Linux.
- The installation files can be downloaded from <https://cran.r-project.org/>.
- Users can install different versions of R in a same machine or computer.
- There is no need to uninstall if you want to upgrade the currently installed R.

<img>

### Starting R

Double click on R icon and you should get this

<img>

You should see an R console.

### Installation of RStudio
RStudio installation files can be downloaded from <http://www.rstudio.com/>.
First, make sure you have RStudio successfully installed.

<img>

#### Starting RStudio
You can double click on RStudio icon and you will see this:

<img>

#### Why RStudio?
- Working with R console is alright. 
- But for many people, they prefer to communicate with R using a graphical user interface (GUI).
- RStudio is the popular GUI and intergrated developement environment (IDE) for R.
- Other R IDE includes Microsoft R

Check this links for more info:

1.  RStudio <https://www.rstudio.com/>
2.  Microsoft R <http://blog.revolutionanalytics.com/2016/01/microsoft-r-open.html>

#### RStudio interface
You should be able to see 4 panes in the layout. You should see that

1. Console - the lower left pane. It tells you about your R information.
2. Source - the upper left pane. It shows the active files.
3. Environment and History - the upper right pane. It shows the currently loaded data files and values, and command history.
4. Miscellaneous - the lower right pane. It contains most important tabs, which are Files, Plots, Packages, Help and Viewer. It list file names, show plots, show packages, display help document and view outputs.

<img>


## Working with packages

### About packages 
R uses packages to perform its tasks.

There are two common packages:

1.  `base` packages 
2.  `user-contributed` packages

- The base packages come with the installation of R
- The base package provides basic but adequate functions to perform many standard data management, visualization and analysis.
- However, user needs to install user-contributed packages if they need to perform functions (tasks) not available in the base package
- User-contributed packages allow users to perform more advanced and more complicated functions 
- There are more than 10200 packages as of March 2017

For a complete list of packages, see <https://cran.r-project.org/web/packages/>

### Package installation
You can install user-contributed packages through:

1.  Internet (to cran)
2.  Github packages
3.  Local zip files

We will learn to install a few small packages.

Basically, a function to install a package will look like this
```{r, eval=FALSE}
install.packages("package.name")
```

To install a package, saya `car`
1.  put your cursor in the CONSOLE pane
2.  type the codes below
```{r, eval=FALSE}
install.packages("car")
```
3.  press Ctrl + ENTER

### Loading packages
Basically, to utilize a package, it has to be loaded using `library()` function,
```{r, eval=FALSE}
library("package.name")
```

For example, we load the newly installed `car` package
```{r, eval=FALSE}
library("haven")
```


## Working directory
In general, R reads and saves data and other files into a working directory. Therefore, a user must create or specify the working directory to work with R. This is a good practice. 

A working directory:

1.  stores all the outputs such as the plots, html files, pdf files 
2.  contains your data

Creating a working directory is a simple BUT an important step. 

Unfortunately, many users do not pay attention to this and forget to set it. So, remember, this is a very important step to work in R.

### Setting a working directory

To set your working directory:

1.  Go back to RStudio's Miscellaneous pane.
2.  In the Files tab, click *...* 
3.  Navigate to the folder containing your data or any folder you want to work in.
4.  Click *More* 
5.  Click *Set as working directory*

or simply use `setwd` function to do so. 
```{r, eval=FALSE}
setwd("path to your folder")
```
for example in Windows
```{r, eval=FALSE}
setwd("C:/myfolder")
```
or in Mac OS/Linux
```{r, eval=FALSE}
setwd("~/myfolder")
```


## Data management
This section is concerned with reading data from dataset and displaying data.

### Reading data set
Easiest is to read .csv file,
```{r, eval=FALSE}
read.csv("file.name")
```

For SPSS file, need `foreign` package,
```{r, eval=FALSE}
library("foreign")
read.spss("file.name")
```

Can read data in table format from text file.
From text file,
```{r, eval=FALSE}
read.table("file.name", header = TRUE)
```

### Viewing data set
Easy, just type the name,
```{r, eval=FALSE}
data
```

Nicer, using `View()`
```{r, eval=FALSE}
View(data)
```

Important tasks
```{r, eval=FALSE}
dim(data)
str(data)
names(data)
```

## More about data management
In this section, we will deal with more advanced data management (subsetting, recoding and creating new variables) and direct data entry (especially useful for tables).

### Subsetting


### Recoding


### Creating new variables


### Direct data entry
