---
title: "Data Management I"
author: "Kamarul Imran Musa"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Data management

This section is concerned with reading data from dataset and displaying data.

There are a few packages to help with reading data, for example

1.  `haven` package
2.  `foreign` package
3.  `readxl` package
4.  `data.table` package
5.  `utils` package

`utils` package come together with R, so you do not need to install it before using it. `haven` package comes with RStudio; to use it, you click the **Import Dataset** menu in RStudio. The rest of the packages needs to be installed into your R library directory. 


## Reading data set

### CSV format

Your datasets may come in the *comma separated value* format or `.CSV`. To read `.CSV` files you can easily use `read.csv()` function. For example, 

```{r}
mycsv.data <- read.csv("cholest.csv")
```

## SPSS format

SPSS dataset comes with `.sav` format. For SPSS, we can use `haven` or `foreign` package to read the data and store it as an object in R. For example

```{r, eval=FALSE}
library("foreign")
myspss.data <- read.spss("cholest.sav", to.data.frame = TRUE)
```


```{r, eval=FALSE}
library("haven")
myspss2.data <- read_sav("cholest.sav")
```

Let us see the difference between the objects created by **foreign** and **haven** packages.

The first one is in the class of the objects. **foreign** package produces object of class data frame, and **haven** produces an object of `tibble`. 

```{r}
class(myspss.data) ; class(myspss2.data)
```

```{r}
head(myspss.data)
```

When we run `summary()` function, we will get this

```{r}
summary(myspss.data)
```

**haven** package does not convert the categorical variable to a factor. It does that to improve efficiency of R.  

```{r}
head(myspss2.data)
```

So, when we run `summarry()` function, we will get this

```{r}
summary(myspss2.data)
```

If we do further checking, we can see that the categorical variables have been converted to *labelled* numerical variables. And that they are in *character chr* format. 

```{r}
str(myspss2.data)
```

if you a to create a factor variable from the character variable, then you can choose one of the following functions below. Each one does something differently.  

```{r}
myspss2.data$sex2 <- as_factor(myspss2.data$sex, "values")
myspss2.data$sex3 <- as_factor(myspss2.data$sex, "labels")
myspss2.data$sex4 <- as_factor(myspss2.data$sex, "values")
myspss2.data$sex5 <- as_factor(myspss2.data$sex, "both")
summary(myspss2.data)
```


## Stata format

```{r}
data <- read.dta("cholest.dta", convert.factors = TRUE)
```

For Excel file, we need `readxl` package,
```{r eval=FALSE}
library("readxl")
data <- read_excel("cholest.xlsx", sheet = 1)
```

### Viewing data set
Easy, just type the name,
```{r, eval=FALSE}
data
```

Nicer, using `View()`
```{r, eval=FALSE}
View(data)
```

View only the first six observations,
```{r}
head(data)
```
and the last six observations,
```{r}
tail(data)
```

Important tasks
```{r, eval=FALSE}
dim(data)
str(data)
names(data)
```

### Exporting data set from R

You can also export data into various formats using similar packages.

For example, 

1.  to export data into a *comma separated version* (.csv) file, we can use `write.csv` function. 
2.  to export data into stata format, we can use `write.dta` function

```{r, eval=FALSE}
write.csv(data, 'data.csv')
write.dta(data, 'data.dta') 
```

## More about data management
In this section, we will deal with more advanced data management (subsetting, recoding and creating new variables) and direct data entry (especially useful for tables).

Let say we use `cholest.csv`,
```{r}
data <- read.csv("cholest.csv")
dim(data)
str(data)
names(data)
head(data); tail(data)
```

### Subsetting

Subsetting means "selecting parts of data". It allows selecting only a number of variables (columns) or observations (rows) from a dataframe. There are ways to do that.

#### Selecting a column (variable) or a row (observation)

Let say, to select `age`
```{r}
data$age
```
to select the 7th observation,
```{r}
data[7, ]
```

#### Selecting columns

Let us create a new data frame with only `chol`, `age` and `sex` as the variables
```{r}
data_col <- subset(data, select = c("chol", "age", "sex"))
str(data_col)
```

alternatively, we can use square brackets
```{r}
data_col <- data[ , c("chol", "age", "sex")]
str(data_col)
```

you may even select by the column numbers
```{r}
data_col <- data[ , c(1:2, 4)]
str(data_col)
```
selecting column 1 to 2, and column 4.

#### Selecting rows

To select 7th to 14th observations,
```{r}
data_row <- data[7:14, ]
data_row
```

Practically, we want to choose observations based on certain criteria, for example those aged > 35 year old,
```{r}
data_row <- subset(data, age > 35)
str(data_row)
```

alternatively, we can use square brackets,
```{r}
data_row <- data[data$age > 35, ]
str(data_row)
```

#### Select rows and columns together

Select those aged > 35, and `chol`, `age`, `sex` variables,
```{r}
data_rc <- subset(data, age > 35 & sex == 1, select = c("chol", "age", "sex"))
str(data_rc)
```

#### Creating a new variable

For example, create age in months,
```{r}
data$age_month <- data$age * 12
data$age_month
```

### Recoding

### Categorize into new variables

#### From a numerical variable

```{r}
data$age_cat <- cut(data$age, breaks = c(-Inf,40,50,Inf),
                    labels = c("< 40", "40-49", "> 50"))
table(data$age_cat)
str(data$age_cat)
```

#### From a categorical variable

Using `age_cat` variable,
```{r}
levels(data$age_cat)
table(data$age_cat)
```
Only one observation labeled as `> 50`. We want to combine `40-49` with `> 50`.
```{r}
library(car)
data$age_cat1 <- recode(data$age_cat, "c('40-49','> 50') = '40 & above'")
table(data$age_cat1)  # combined
```

### Direct data entry

We may also enter short data directly using `read.table`.

For example, a standard data frame,

ID | Group | BMI |
---|---|---:|
1 | Fat | 30 |
2 | Fat | 31 | 
3 | Fat | 32 |
4 | Thin | 20 |
5 | Thin | 19 |
6 | Thin | 18 |

```{r}
data_frame <- read.table(header = TRUE, text = "
ID Group BMI
1 Fat 30
2 Fat 31
3 Fat 32
4 Thin 20
5 Thin 19
6 Thin 18
")
str(data_frame)
data_frame
```

or a table,

| | Cancer | No Cancer
| --- | ---: | ---:
| Smoker | 80 | 10
| Non-smoker | 5 | 100

```{r}
data_table <- read.table(header = FALSE, text = "
80 10
5 100
                        ")
colnames(data_table) <- cancer <- c("Cancer", "No Cancer")
rownames(data_table) <- c("Smoker", "Non-smoker")
str(data_table)  # still a data frame, but laid out in form of a table.
data_table
```
The numbers are separated by space. We set the row and column names by `rownames` and `colnames` respectively.

## Summary
In this chapter, we learned some basics in using R effectively. In the next chapter, we are going to learn about how to explore the variables by means of basic descriptive statistics.
