---
output:
  pdf_document: default
---

# Descriptive statistics
In this chapter, we will learn how to obtain a number of important descriptive statistics. The statistics will be obtained based on the variable types and groups. We will also learn how to perform cross-tabulation.

## One variable
We will start by using `chickwts` dataset that contains both numerical (`weight`) and categorical (`feed`) variables. We view the first six observations,
```{r}
head(chickwts)
```
the last six observations.
```{r}
tail(chickwts)
```

Next, view the details of the data,
```{r}
str(chickwts)
```
Here we have 71 rows (71 subjects) and two columns (two variables). `weight` is a numerical variable and `feed` is a factor, i.e. a categorical variable. `feed` consists of six categories or levels.

We can view the levels in `feed`,
```{r}
levels(chickwts$feed)
```

### A numerical variable
A numberical variable is described by a number of descriptive statistics below.

To judge the central tendency of the `weight` variable, we obtain its mean,
```{r}
mean(chickwts$weight)
```
and median,
```{r}
median(chickwts$weight)
```
To judge its spread and variability, we can view its minimum, maximum and range
```{r}
min(chickwts$weight)
max(chickwts$weight)
range(chickwts$weight)
```
and obtain its standard deviation (SD)
```{r}
sd(chickwts$weight)
```
variance,
```{r}
var(chickwts$weight)
```
quantile,
```{r}
quantile(chickwts$weight)
```
and interquartile range (IQR)
```{r}
IQR(chickwts$weight)
```
There are nine types of quantile algorithms in R (for `quantile()` and `IQR`), t()he default being type 7. You may change this to type 6 (Minitab and SPSS),
```{r}
quantile(chickwts$weight, type = 6)
IQR(chickwts$weight, type = 6)
```
In addition to SD and IQR, we can obtain its median absolute deviation (MAD),
```{r}
mad(chickwts$weight)
```
It is actually simpler to obtain most these in a single command,
```{r}
summary(chickwts$weight)
```
even simpler, obtain all of the statistics using `describe()` in the `psych` package
```{r eval = FALSE}
install.packages("psych")
```
```{r, message=FALSE, error=FALSE}
library(psych)
describe(chickwts$weight)
```

### A categorical variable
A categorical variable is described by its count, proportion and percentage by categories.

We obtain the count of the `feed` variable,
```{r}
summary(chickwts$feed)
table(chickwts$feed)
```
both `summary()` and `table()` give the same result.

`prop.table` gives the proportion of the result from the count.
```{r}
prop.table(table(chickwts$feed))
```
the result can be easily turned into percentage,
```{r}
prop.table(table(chickwts$feed))*100
```
To view the count and the percentage together, we can use `cbind`,
```{r}
cbind(n = table(chickwts$feed), 
      "%" = prop.table(table(chickwts$feed))*100)
```
We need the quotation marks " " around the percentage sign %, because % also serves as a mathematical operator in R.

## Two variables and more
Just now, we viewed all the statistics as applied to a variable. In this part, we are going to view the statistics on a number of variables. This includes viewing a group of numerical variables or categorical variables, or a mixture of numerical and categorical variables. This is relevant in a sense that, most of the time, we want to view everything in one go (e.g. the statistics of all items in a questionnaire), compare the means of several groups and obtain cross-tabulation of categorical variables.

### Numerical variables
Let us use `women` dataset and explore the dataset,
```{r}
head(women)
tail(women)
str(women)
```
which consists of `weight` and `height` numerical variables.

The variables can be easily viewed together by `summary`,
```{r}
summary(women)
```
even better using `describe` (`psych` package),
```{r}
library(psych)
describe(women)
```

### Categorical variables
Let us use `infert` dataset,
```{r}
head(infert)
str(infert)
```
We notice that `induced`, `case` and `spontaneous` are not yet set as categorical variables, thus we need to `factor` the variables. We view the value labels in the dataset description,
```{r}
?infert
```
We label the values in the variables according to the description as
```{r}
infert$induced <- factor(infert$induced, levels = 0:2, 
                         labels = c("0", "1", "2 or more"))
infert$case <- factor(infert$case, levels = 0:1, 
                      labels = c("control", "case"))
infert$spontaneous <- factor(infert$spontaneous, 
                             levels = 0:2, 
                             labels = c("0", "1", "2 or more"))
str(infert)
```
and we now all these variables are turned into factors.

Again, the variables can be easily viewed together by `summary()`,
```{r}
summary(infert[c("education", "induced", "case", "spontaneous")])
```
We do not use `table()` here in form of `table(infert[c("education", "induced", "case", "spontaneous")])` because `table()` used in this form will give us 3-way cross-tabulation instead of count per categories. Cross-tabulation of categorical variables will be covered later.

To obtain the proportion and percentage results, we have to use `lapply()`,
```{r}
lapply(infert[c("education", "induced", "case", "spontaneous")], 
       function(x) summary(x)/length(x))
lapply(infert[c("education", "induced", "case", "spontaneous")], 
       function(x) summary(x)/length(x)*100)
```
because we need `lappy()` to obtain the values for each of the variables. `lapply()` goes through each variable and performs this particular part,
```{r eval = FALSE}
function(x) summary(x)/length(x)
```
`function(x)` is needed to specify some extra operations to any basic function in R, in our case `summary(x)` divided by `length(x)`, in which the summary results (the counts) are divided by the number of subjects (`length(x)` gives us the "length" of the variable).

Now, since we already learned about `lapply()`, we may also obtain the same results by using `summary()` (within `lapply()`), `table()` and `prop.table()`.
```{r}
lapply(infert[c("education", "induced", "case", "spontaneous")], 
       summary)
lapply(infert[c("education", "induced", "case", "spontaneous")], 
       table)
lapply(infert[c("education", "induced", "case", "spontaneous")], 
       function(x) prop.table(table(x)))
lapply(infert[c("education", "induced", "case", "spontaneous")], 
       function(x) prop.table(table(x))*100)
```
Notice here, whenever we do not need to specify extra operations on a basic function, e.g. `summary()` and `table()`, all we need to write after the comma in `lapply()` is the basic function without `function(x)` and `(x)`.

## Groups and cross-tabulations
We intentionally went through the descriptive statistics of a variable, followed by a number of variables of the same type. This will give you the basics in dealing with the variables. Most commonly, the variables are described by groups or in form cross-tabulated counts/percentages.

### By groups
To obtain all the descriptive statistics by group, we can use `by` with the relevant functions. Let say we want to obtain the statistics by case and control (`case`). We start with numerical variables
```{r}
by(infert[c("age", "parity")], infert$case, summary)
by(infert[c("age", "parity")], infert$case, describe)
```
We can also use `describeBy()`, which is an the extension of `describe()` in the `psych` package.
```{r}
describeBy(infert[c("age", "parity")], group = infert$case)
```
which gives us an identical result.

If you want to obtain results using the basic functions (i.e. `mean()`, `median()`, `quantile()`, `IQR()` and `mad()`), you need to use `lappy()` within `by()`, because they could not handle many variables, for example for `mean()` and `IQR()`,
```{r}
by(infert[c("age", "parity")], infert$case, 
   function(x) lapply(x, mean))
by(infert[c("age", "parity")], infert$case, 
   function(x) lapply(x, IQR))
```

For categorical variables, using `summary()`
```{r}
by(infert[c("education", "induced", "spontaneous")], infert$case, 
   summary)
by(infert[c("education", "induced", "spontaneous")], infert$case, 
   function(x) lapply(x, function(x) summary(x)/length(x)))
by(infert[c("education", "induced", "spontaneous")], infert$case, 
   function(x) lapply(x, function(x) summary(x)/length(x)*100))
```
or by using `table()`
```{r}
by(infert[c("education", "induced", "spontaneous")], infert$case, 
   function(x) lapply(x, table))
by(infert[c("education", "induced", "spontaneous")], infert$case, 
   function(x) lapply(x, function(x) prop.table(table(x))))
by(infert[c("education", "induced", "spontaneous")], infert$case, 
   function(x) lapply(x, function(x) prop.table(table(x))*100))
```
Please note that simply replacing `table()` for `summary()` as in `by(infert[c("education", "induced", "spontaneous")], infert$case, table)` will not work as intended. `education` will be nested in `induced`, which is nested in `spontaneous`, listed by `case` instead. And yes, to obtain the proportions and percentages, it gets slightly more complicated as we have to specify `function()` twice in `by()`.

### Cross-tabulation
As long as the categorical variables are already `factor`ed properly, there should not be a problem to obtain the cross-tabulation tables. For example between `education` and `case`,
```{r}
table(infert$education, infert$case)
```
We may also include row and column headers, just like `cbind`,
```{r}
table(education = infert$education, case = infert$case)
```
Since we are familiar with the powerful `lappy`, we can use it to get cross-tabulation of all of the factors with `case` status,
```{r}
lapply(infert[c("education", "induced", "spontaneous")], 
       function(x) table(x, infert$case))
```
We may also view subgroup counts (nesting). Here, the cross-tabulation of `education` and `case` is nested within `induced`
```{r}
table(infert$education, infert$case, infert$induced)
```
which will look nicer if we apply `by`
```{r}
by(infert[c("education", "case")], infert$induced, table)
```

## Customizing text outputs
Text outputs will look nicer by combining every bits and parts of the outputs into custom-made texts and tables. There are a number of ways to achieve this. We will do this by utilizing basic functions.

We again use `cholest.sav` from previous chapter. Load the dataset as `cholest`,
```{r}
library(foreign)
cholest <- read.spss("cholest.sav", to.data.frame = TRUE)
```
```{r eval=FALSE}
str(data)
```
```
## 'data.frame':	80 obs. of  5 variables:
##  $ chol    : num  6.5 6.6 6.8 6.8 6.9 7 7 7.2 7.2 7.2 ...
##  $ age     : num  38 35 39 36 31 38 33 36 40 34 ...
##  $ exercise: num  6 5 6 5 4 4 5 5 4 6 ...
##  $ sex     : Factor w/ 2 levels "female","male": 2 2 2 2 2 2 2 2 2 2 ...
##  $ categ   : Factor w/ 3 levels "Grp A","Grp B",..: 1 1 1 1 1 1 1 1 1 1 ...
##  - attr(*, "variable.labels")= Named chr  "cholesterol in mmol/L" "age in year"
##  "duration of exercise (hours/week)" "" ...
##  ..- attr(*, "names")= chr  "chol" "age" "exercise" "sex" ...
##  - attr(*, "codepage")= int 65001
```

### `cbind` and `rbind`
We were introduced to `cbind()` earlier in this chapter. We will further use `cbind()` to customize our outputs. In addition, we will use its sibling, `rbind()`.

Let say we want to view mean, standard deviation (SD) and sample size (n) together,
```{r}
mean(cholest$age)
sd(cholest$age)
length(cholest$age)
```

First utilize the basic `cbind()`,
```{r}
cbind(mean = mean(cholest$age), sd = sd(cholest$age), 
      n = length(cholest$age))
```

and then we can give it a proper row name,
```{r}
chol_c <- cbind(mean = mean(cholest$age), sd = sd(cholest$age), 
                n = length(cholest$age))
rownames(chol_c) <- "Cholestrol"
chol_c
```

Compare `cbind()` with `rbind()`. `rbind()` combines the values by row, while `cbind()` combines the values by column. Thus, you can customize the outputs based on your preference.
```{r}
rbind(mean = mean(cholest$age), sd = sd(cholest$age), 
      n = length(cholest$age))
chol_r = rbind(mean = mean(cholest$age), sd = sd(cholest$age), 
               n = length(cholest$age))
colnames(chol_r) <- "Cholestrol"
chol_r
```

Now we can add in `lappy()` to come up with vectors of mean and SD for the selected variables. `cbind()` and `rbind()` can also combine vectors,
```{r}
mean_cholest <- lapply(cholest[, c("chol", "age", "exercise")], mean)
sd_cholest <- lapply(cholest[, c("chol", "age", "exercise")], sd)
cbind(mean = mean_cholest, SD = sd_cholest, 
      n = lengths(cholest[, c("chol", "age", "exercise")]))
rbind(mean = mean_cholest, SD = sd_cholest, 
      n = lengths(cholest[, c("chol", "age", "exercise")]))
```

Now, we can edit the variable names to make the results more presentable,
```{r}
names(mean_cholest) <- c("Cholestrol", "Age", "Exercise")
cbind(mean = mean_cholest, SD = sd_cholest, 
      n = lengths(cholest[, c("chol", "age", "exercise")]))
rbind(mean = mean_cholest, SD = sd_cholest, 
      n = lengths(cholest[, c("chol", "age", "exercise")]))
```

Now, let us try `cbind()` and `rbind()` on categorical variables, `sex` and `categ`,
```{r}
count_cholest <- sapply(cholest[c("sex", "categ")], summary)
count_cholest
perc_cholest <- sapply(cholest[c("sex", "categ")], function(x) summary(x)/length(x)*100)
perc_cholest
```

then we list down by variables,
```{r}
list(Sex = cbind(n = count_cholest$sex, "%" = perc_cholest$sex),
     Category = cbind(n = count_cholest$categ, 
                      "%" = perc_cholest$categ))
```

### `data.frame()` and `matrix()`
These two functions work like `cbind()`. They are very handy to present results to look like a nice table.

Using `data.frame()`,
```{r}
data.frame(mean = mean_cholest, SD = sd_cholest, 
      n = lengths(cholest[, c("chol", "age", "exercise")]))
```

Using `matrix()`,
```{r}
matrix(c(mean_cholest, sd_cholest, 
         n = lengths(cholest[, c("chol", "age", "exercise")])), 
       nrow = 3, ncol = 3,
       dimnames = list(names(cholest[, c("chol", "age", "exercise")]),
                       c("mean", "SD", "n")))
```

### `paste0()`
We can also use the `table()` and `paste0()` as follows,
```{r}
tab_categ = table(Category = cholest$categ)
per_categ = prop.table(tab_categ)*100
cell_categ = paste0(tab_categ, " (", per_categ, "%)")
tab_per_categ = tab_categ  # just to set the dimension of `tab_per_categ`
tab_per_categ[] = cell_categ[]
tab_per_categ
```

In another example for cross-tabulation,
```{r}
tab = table(Category = cholest$categ, Gender = cholest$sex); tab  # count
per = prop.table(table(Category = cholest$categ, Gender = cholest$sex))*100
per  # % 
cbind(tab, per)
addmargins(tab)  # marginal counts
# nicer view
cell = paste0(tab, " (", per, "%)")
str(tab)
tab1 = tab
tab1[] = cell[]
tab1
ftable(tab1)  # nicer 'flat' view
```

### `cat()`
Lastly, `cat()` can be used to write combine relevant outputs in text format.
```{r}
cat("For cholestrol, the mean was ", round(mean(cholest$chol), 2), 
    " (SD = ", round(sd(cholest$chol)), ") in a sample of ",
    length(cholest$chol), " subjects.", sep = "")
```

## Summary
In this chapter, we learned about how to handle numerical and categorical variables and obtain the basic and relevant descriptive statistics. We also learned how to combine outputs into custom made tables and texts.

In the next chapter, we are going to learn about how to explore the variables visually in form of the relevant graphs and plots.
